PanGraph: scalable bacterial pan-genome graph construction

### 2022-09-29 main.tex
started panx section in main

\graphicspath{{../figs/}{./figs}}

@@

@@

In order to assess the accuracy of the pangenome graph inferred by PanGraph, and to quantify its performance characteristics as a function of input size, we generated synthetic data.
Specifically, we simulated populations of size $N=100$ and sequence length $L=50000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (each with independent rates set to $10^{-2}$ per generation), and horizontal sequence transfer that occurs with tunable rate $r$ per generation per genome.
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line toolsuite for external use.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
    \caption{{\bf Accuracy against synthetic data}.
        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} and \textit{asm20} option, and \textit{mmseqs2}. In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp. The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of 100bp, as a function of average pairwise sequence divergence. Line and shaded area represents mean and standard deviation over 25 repetitions.
        Graph reconstruction with \textit{minimap2} and \textit{asm20} is accurate up to around 5\% average sequence divergence, while for \textit{mmseqs2} accuracy is maintained until $\sim$12\% divergence.
    }
    \label{fig:toy-accuracy}
\end{figure}

We quantified the accuracy of PanGraph algorithm by computing the distance, $\delta$, of the inferred breakpoints relative to their known locus stored from the evolutionary simulation.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $r$ and thus underlying graph complexity.
The predominant determinant of accuracy was the sample diversity, shown in Fig. \ref{fig:toy-accuracy} which displays the distributions of $\delta$ for $100$ independent simulations ran for each mutation rate $\mu$.
For approximately clonal populations, $\sim 80\%$ of breakpoints were estimated to be within $10$bp away from the ``true" location.
Conversely, for samples with average pairwise diversity above $2\%$, $\sim 80\%$ of breakpoints are estimated only within $100$ bp.
This limitation is primarily driven by lack of sensitivity in the pairwise alignment phase performed using \emph{minimap2} \cite{li2018minimap2}.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
    \caption{{\bf Algorithm performance}.
        PanGraph scales linearly with the number of input genomes.
        This is a direct result of the guide tree simplification.
        The solid line and ribbons display the mean and standard deviation respectively.
        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
    }
    \label{fig:toy-performance}
\end{figure}

The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length.
The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of Minimap2 \cite{li2018minimap2}.

@@

We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}.
Obviously the ``true" breakpoints are not known \emph{a priori} and thus the direct analog of Fig. \ref{fig:toy-accuracy} is not possible.
However, we postulate that most \emph{bona fide} breakpoints generated from evolutionary events will happen at or near gene boundaries; there is preliminary evidence for this assumption \cite{oliveira2017chromosomal}.
We compared the empirical distribution of distances to gene boundaries to a null distribution generated from the data, in which breakpoints are sampled uniformly from each genome, constrained to produce equivalent number of partitions as was measured by PanGraph.
For all tested cases, we find our boundaries cluster closer to genes than would be expected by chance.
The median distance from a Pangraph boundary to annotated gene ranges between 50-100 bp, depending on the species, while the random breakpoint is, on average, 600 bp off.

@@

The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.

\begin{table}[hb]
    \caption{{\bf Real-world Performance}.
        Elapsed time and maximum RAM usage needed to compute the pangraph for the following collections of RefSeq genomes.
        All runs were performed on an AMD Ryzen 7 2700X utilizing 8 cores.
    }
    \begin{tabular}{l r r r }
        \hline\hline
        Species         & Sample size & Wall time (min) & Max memory (GB) \\
        \hline
        E.~coli         & 307         & 426             & 11.7            \\
        H.~pylori       & 85          & 102             & 1.2             \\
        K.~pneumoniae   & 109         & 132             & 2.1             \\
        M.~tuberculosis & 51          & 91              & 0.4             \\
        P.~marinus      & 10          & 12              & 0.1             \\
        \hline
    \end{tabular}
    \label{table:panx-performance}
\end{table}

\begin{figure*}[htb]
    \includegraphics[width=.8\textwidth]{figs/proj_single_escherichia_coli.pdf}
    \caption{{\bf Test of graph marginalization}.}
    \label{fig:marginalization}
\end{figure*}

>>>>

\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\Lcore}{L_\text{core}}
\newcommand{\Lpang}{L_\text{pang}}
\newcommand{\Lgen}{L_\text{gen}}
\newcommand{\dcore}{\langle d_\text{core} \rangle}

@@

% TODO: check consistency of numberings of figures and sections between SI and main text
% TODO: real data
% TODO: - dataset
% TODO:     - table
% TODO: - size
% TODO:     - figure
% TODO: - benchmark
% TODO:     - figure
% TODO: - marginalization
% TODO: 
% TODO: Table: leave all the columns? Improve description? Depending on what is useful afterwards
% TODO: SI Section IIA: superfluous?
% TODO: minimap2 asm10 superfluous?

@@

As a first validation step we use generated synthetic data to quantify the performance characteristics of \textit{PanGraph} as a function of input size, and its accuracy as a function of sequence diversity.

\paragraph*{Generation of synthetic data}

We simulated populations of size $N=100$ and sequence length $L=50\,000$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI Section I for details).
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line toolsuite for external use.

\paragraph*{Performances as a function of dataset size}

The algorithmic complexity was measured empirically by simulating populations of increasing size and average sequence length, and generating pangenome graphs for the resulting sequences.
The mean and standard deviation of results obtained from 50 iterations are shown in Fig. \ref{fig:toy-performance}.
Importantly, PanGraph's computational complexity grows \emph{linearly} with the number of input genomes.
We note that the PanGraph scales approximately log-linear with average sequence length, as expected from the underlying algorithmic complexity of \textit{minimap2} \cite{li2018minimap2}.
Benchamrks of CPU and memory usage are reported in SI Section IIA.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{figs/benchmark.pdf}
    \caption{{\bf Algorithm performance}.
        PanGraph scales linearly with the number of input genomes.
        This is a direct result of the guide tree simplification.
        The solid line and ribbons display the mean and standard deviation over 50 runs.
        All runs were performed utilizing 8 cores.
    }
    \label{fig:toy-performance}
\end{figure}

\paragraph*{Accuracy as a function of sequence divergence}

We quantified the accuracy of PanGraph algorithm by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf. SI Section IIB for details). We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and HGT $h$. For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel, namely \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} option and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal gene transfer $h$ and thus underlying graph complexity. The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf. SI Fig.~2). For low-diversity samples breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf. SI Fig.~3). The choice of alignment kernel influences the threshold diversity at which accuracy is lost.\\
For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default precision threshold for Pangraph of 100 bp, displayed in Fig.~\ref{fig:toy-accuracy}. On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at a threshold pairwise divergence of around 5\%, while \textit{mmseqs2} is accurate up to average sequence divergence of around 12\%, at the cost of higher computational time.

\begin{figure}[htb]
    \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
    \caption{{\bf Accuracy against synthetic data}.
        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels :\textit{minimap2} with \textit{asm10} and \textit{asm20} option, and \textit{mmseqs2}. In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp. The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of 100bp, as a function of average pairwise sequence divergence. Line and shaded area represents mean and standard deviation over 25 repetitions. \textit{mmseqs2} maintains accuracy at higher divergence, at the cost of more computational time.
    }
    \label{fig:toy-accuracy}
\end{figure}

@@

We additionally validated PanGraph on natural populations sampled from RefSeq \cite{o2016reference}, focusing on the properties of the resulting pangenome graphs for datasets of different size and diversity.

\paragraph*{Dataset characterization}

We downloaded from RefSeq \cite{o2016reference} completely assembled chromosomes from 5 different bacterial species: \textit{Escherichia Coli} (EC), \textit{Klebsiella Pneumoniae} (KP), \textit{Helicobacter Pylori} (HP), \textit{Mycobacterium Tuberculosis} (MT) and \textit{Prochlorococcus Marinus} (PM). These data had been previously analyzed using PanX \cite{ding2018panx}, which allowed us to estimate the size of the pangenome and core genome, and the average pairwise divergence on core genes (cf. Table~\ref{table:panx-dataset} and SI Section~IIIA).
% The \textit{E.~Coli} (EC) dataset is the one with the highest number of isolates (307 chromosomes) and the largest pangenome. The species with the highest core genome divergence are \textit{H.~Pylori} (HP) and \textit{P. Marinus} (PM). For the former, the divergce sits at the limit of what the \textit{minimap2} alignment kernel is able to accurately merge (cf. Fig.~\ref{fig:toy-accuracy}), while for the latter the diversity is beyond even the limits of \textit{mmseqs2}.

\begin{table}[h]
    \setlength{\tabcolsep}{6pt}
    \begin{tabular}{l c c c c c }
        \hline\hline
        Species         & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\dcore$ \\
        \hline
        E.~Coli         & 307 & 5.0           & 17.7     & 0.7      & 1.6\%    \\
        K.~Pneumoniae   & 109 & 5.3           & 13.0     & 2.3      & 0.8\%    \\
        H.~Pylori       & 85  & 1.6           & 1.9      & 0.6      & 4.2\%    \\
        M.~Tuberculosis & 51  & 4.4           & 4.1      & 2.4      & 0.03\%   \\
        P.~Marinus      & 10  & 1.8           & 3.3      & 0.7      & 26.9\%   \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset properties}. Columns represent:
        (1) number of isolates,
        (2) average chromosome lenght in Mbps,
        (3) total pangenome legnth in Mbp,
        (4) total core genome length in Mbp,
        (5) average pairwise divergence of core genes.
    }
    \label{table:panx-dataset}
\end{table}

\paragraph*{Benchmark}

Using \textit{PanGraph}, we build multiple pangenome graphs for each species in the dataset. We test all the different alignment kernels (\textit{minimap2} with \textit{asm10} or \textit{asm20} option, or \textit{mmseqs2}). Moreover we also vary the value of the pseudo-energy parameters $\alpha,\beta$ from eq.~(\ref{eq:pseudo-energy}), using either the standard values ($\alpha=100$, $\beta=10$) or null values ($\alpha = \beta = 0$).\\
\textit{PanGraph} can build pangenome graphs comprising hundreds of isolates in a few hours (5-6h for 307 EC isolates with \textit{minimap2} kernel, cf. Fig~\ref{fig:panx-benchmark}A). Use of \textit{mmseqs2} consistently requires longer time, but provides higher accuracy when merging highly diverged sequences. This is the case for the HP dataset, whose core genome divergence sits at the limit of what the \textit{minimap2} kernel can accurately align. In this case the use of \textit{mmseqs2} improves both sequence compression and core pangenome fraction (cf. Fig~\ref{fig:panx-benchmark}B-C).\\
% beta poses a limit to the divergence. Removing it..
% but increases fragmentation (cite supplementary)
% talk about PM
% average compression
% core genome fraction compatible?

\paragraph*{Scaling with dataset size}

@@

\section{Graph marginalization}

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
    \caption{{\bf Test of graph marginalization}.}
    \label{fig:marginalization}
\end{figure*}

% % TODO: talk more extensively about the meaning of the results?
% \begin{figure*}[htb]
%     \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
%     \caption{{\bf Benchmark on real data}.
%         We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species. For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters (standard or null values).
%         \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
%         \textbf{B}: fraction of core pangenome in the pangenome graph.
%         \textbf{C}: sequence compression, measured as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph.
%     }
%     \label{fig:panx-benchmark}
% \end{figure*}

% The computational time and maximum memory required to build each alignment is summarized in Table \ref{table:panx-performance}.

% \begin{figure*}[htb]
%     \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
%     \caption{{\bf Test of graph marginalization}.}
%     \label{fig:marginalization}
% \end{figure*}
